### 2014-08-04 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: update input files, update report from comments

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report.  The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas.  The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door.  After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at face height.  Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. The captain was overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure but later succumbed to his injuries.

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of all of these studies has been to improve the safety of fire fighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using Fire Dynamics Simulator and Smokeview to provide insight into the fire development and thermal conditions that might have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding times are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], preceded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgment from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2\textsuperscript{nd} floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the [captain] and other crew member untangled. The TL39 Lieutenant and SQ5 fire fighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NIOSH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & First floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~(\ref{fig:simp_geom}). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~(\ref{fig:alpha_ex}) shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

@@

An unknown parameter in Table~\ref{tab:vents} that needed to be estimated was the time when the second floor interior/porch door failed (2~min and 40~s into the simulation). The area of failure was also estimated based on the following observations. The bottom hinge from the door appeared to be closed throughout the fire due to the lack of fire damage and the lack of soot deposited on the interior faces of the hinge and remaining wood, as shown in Fig.~(\ref{fig:door_frame}). It appears that the door collapsed and failed partially open. In the simulations, only the upper portion of the door was assumed to fail. While this time and area may not be exact, the takeaway from the analysis in the following sections should be impact of the door's failure on the interior conditions of the structure.

@@

For the simulation in this study, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g} }
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the fire as it is specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using 10 cm grid cells, the ratio of the characteristic fire diameter to the cell size ($D^*/\dx$) is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~(\ref{fig:geom_grid}) shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

\section{Heat Release Rate}
\label{HRR}
The heat release rate output data from FDS quantifies the simulated rate of energy release in the simulation. Figure~(\ref{fig:hrr}) shows the prescribed HRR versus time values based on the prescribed fires along with the FDS results for the two ventilation scenarios that were described in Section~\ref{Vents}. The two vertical lines represent the time at which a firefighter opened the back door of the structure ($t$ = 130 s) and the time at which the second floor interior door was estimated to fail ($t$ = 160 s). Due to a lack of information about the time of fire ignition or how fast the fire grew, these parameters needed to be estimated.  The growth rate shown in Fig.~(\ref{fig:hrr}) (10~MW in approximately 20~s) is fast because the incipient stage of the fire growth was not considered in this study. The maximum size of the fire used in this study was selected such that the conditions would become ventilation limited quickly to minimize computational time while still aligning with the timeline observations of video and picture evidence from the fire. The accuracy of heat transfer to the walls of the structure may be sacrificed by using this approach; however, all of the window, door, and wall failures were driven by the event timeline rather than failure temperature criteria.

@@

In Fig.~(\ref{fig:hrr}), the solid line represents the baseline FDS simulation (this follows the noted door openings from the incident). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach of the rear porch wall on the left side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~(\ref{fig:hrr}), the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to combust all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~(\ref{fig:smv_ext_fire}) shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

Figure~(\ref{fig:hrr}) indicates that the HRRs for both of the ventilation scenarios in FDS are identical until 130~s (HRR $\approx$ 7~MW). At 130~s, the rear door is opened in the baseline simulation. At 135~s, there is an additional breach on the left side of the structure (Table~\ref{tab:vents}). The open door and breach act as additional pressure releases for the structure. Again, the abrupt change in ventilation causes the HRR to increase from 7~MW to 9.7~MW at 139~s. At the time of the door failure (160~s), the HRR in both cases was approximately 9~MW because of the increased amount of ventilation. While there might be some locally different burning characteristics between the baseline simulation and the alternative simulation, the interior conditions in both cases presented a significant hazard to firefighters inside of the structure. Therefore, the analysis will focus on the results of the baseline simulation.

@@

As the fire grows in the attic and rear porch, the pressure rises. Figure~(\ref{fig:pres_159s}) shows the pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa over pressure 1~m off the floor to greater than 10~Pa (0.00145~psi) at the ceiling. Gases at an elevated pressure will always move towards a region of lower pressure. The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~(\ref{fig:velo_175s}), 15~s after the interior door fails, confirm this flow direction within the structure.

@@

\section{Temperature}
\label{temp}
Analysis of the temperatures from the FDS simulation will focus on the high hazard areas of the structure: the attic, the rear porch and stairwell, and the interior hallway which connects the rear porch to the front stairwell. The temperature profile at these high hazard areas will be shown via several snapshots in time to illustrate the change in the interior conditions. The first snapshots are shown at a time of 129~s into the simulation, just before a firefighter opened the rear porch door (as shown in Fig.~(\ref{fig:temp_129s})), and 159~s into the simulation, just before the interior door fails (as shown in Fig.~(\ref{fig:temp_159s})).

>>>>

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report and are based on the fee scenario including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\chapter{Introduction}
Part of the function of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  softwareto provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the creation of a flow path inside the structure.  The document is organized in the following fashion:  Section~\ref{fire_sum} describes the events of the fire incident, Section~\ref{model} describes the relevant input parameters and assumptions used to develop the simulations, Section~\ref{results} presents simulation results, and Section~\ref{discussion} discusses the model results as they relate to firefighter safety and effectiveness.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident were documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], proceeded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgment from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2\textsuperscript{nd} floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the [captain] and other crew member untangled. The TL39 Lieutenant and SQ5 firefighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~\ref{fig:simp_geom}.
\begin{table}
\centering
\captionof{table}{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & First floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
The Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}. The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

@@

An unknown parameter in Table~\ref{tab:vents} that needed to be estimated was the time when the second floor interior/porch door failed (2~min and 40~s into the simulation). The area of failure was also estimated based on the following observations. The bottom hinge from the door appeared to be closed throughout the fire due to the lack of fire damage and the lack of soot deposited on the interior faces of the hinge and remaining wood, as shown in Fig.~\ref{fig:door_frame}. It appears that the door collapsed and failed partially open. In the simulations, only the upper portion of the door was assumed to fail. While this time and area may not be exact, the takeaway from the analysis in the following sections should be impact of the door's failure on the interior conditions of the structure.

@@

For the simulation in this study, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g} }
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the fire as it is specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using 10 cm grid cells, the ratio of the characteristic fire diameter to the cell size ($D^*/\dx$) is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

\section{Heat Release Rate}
\label{HRR}
The heat release rate output data from FDS quantifies the simulated rate of energy release in the simulation. Figure~\ref{fig:hrr} shows the prescribed HRR versus time values based on the prescribed fires along with the FDS results for the two ventilation scenarios that were described in Section~\ref{Vents}. The two vertical lines represent the time at which a firefighter opened the back door of the structure ($t$ = 130 s) and the time at which the second floor interior door was estimated to fail ($t$ = 160 s). Due to a lack of information about the time of fire ignition or how fast the fire grew, these parameters needed to be estimated.  The growth rate shown in Fig.~\ref{fig:hrr} (10~MW in approximately 20~s) is fast because the incipient stage of the fire growth was not considered in this study. The maximum size of the fire used in this study was selected such that the conditions would become ventilation limited quickly to minimize computational time while still aligning with the timeline observations of video and picture evidence from the fire. The accuracy of heat transfer to the walls of the structure may be sacrificed by using this approach; however, all of the window, door, and wall failures were driven by the event timeline rather than failure temperature criteria.

@@

In Fig.~\ref{fig:hrr}, the solid line represents the baseline FDS simulation (this follows the noted door openings from the incident). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach of the rear porch wall on the left side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~\ref{fig:hrr}, the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to combust all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~\ref{fig:smv_ext_fire} shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

Figure~\ref{fig:hrr} indicates that the HRRs for both of the ventilation scenarios in FDS are identical until 130~s (HRR $\approx$ 7~MW). At 130~s, the rear door is opened in the baseline simulation. At 135~s, there is an additional breach on the left side of the structure (Table~\ref{tab:vents}). The open door and breach act as additional pressure releases for the structure. Again, the abrupt change in ventilation causes the HRR to increase from 7~MW to 9.7~MW at 139~s. At the time of the door failure (160~s), the HRR in both cases was approximately 9~MW because of the increased amount of ventilation. While there might be some locally different burning characteristics between the baseline simulation and the alternative simulation, the interior conditions in both cases presented a significant hazard to firefighters inside of the structure. Therefore, the analysis will focus on the results of the baseline simulation.

@@

As the fire grows in the attic and rear porch, the pressure rises. Figure~\ref{fig:pres_159s} shows the pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa over pressure 1~m off the floor to greater than 10~Pa (0.00145~psi) at the ceiling. Gases at an elevated pressure will always move towards a region of lower pressure. The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~\ref{fig:velo_175s}, 15~s after the interior door fails, confirm this flow direction within the structure.

@@

\section{Temperature}
\label{temp}
Analysis of the temperatures from the FDS simulation will focus on the high hazard areas of the structure: the attic, the rear porch and stairwell, and the interior hallway which connects the rear porch to the front stairwell. The temperature profile at these high hazard areas will be shown via several snapshots in time to illustrate the change in the interior conditions. The first snapshots are shown at a time of 129~s into the simulation, just before a firefighter opened the rear porch door (as shown in Fig.~\ref{fig:temp_129s}), and 159~s into the simulation, just before the interior door fails (as shown in Fig.~\ref{fig:temp_159s}).
